---
title: "ADA-GeospatialVignette-StevenRubenIzzi"
output: html_document
---
Packages to install:
```{r, message=FALSE}
library(sf)
library(sp)
library(dplyr)
library(ggmap)
library(ggplot2)
```
## 1. Read in geospatial data (GPX)
The {sf} package allows spatial objects to be stored as data frames that are easily manipulated with dplyr. The spatial geometry of a feature is stored in a list-column called geometry, which can be as complex as needed to store more geographic information in a single variable for each feature.

However, the *st_read()* function can read in a number of different filetypes and automatically recognizes file suffixes. Use *st_drivers()* to print all file formats that can be read in by st_read().

In this example, we will be loading in GPX files containing chimpanzee group movement data collected on a handheld GPS tracker by a researcher following on foot the members of two adjacent groups of chimpanzees in Kibale National Park, Uganda.
```{r}
f <- file.choose() # west tracks
f2 <- file.choose() # east/central tracks
```

You can use *st_layers(f)* to print the layer types a file contains, along with the number of features and fields for each.
```{r}
st_layers(f)
```
The layer we are interested in here is "track_points", which we can specify with the *layer* argument when we read in our files with st_read():
```{r}
west_sf <- st_read(f, layer = "track_points")
class(west) # data is loaded in as an "sf" dataframe
west <- as_tibble(west_sf)
# Repeat for east/central group tracks
eastcent_sf <- st_read(f2, layer = "track_points")
eastcent <- as_tibble(eastcent_sf)
```
We can do an inital visualization of our data using the *plot* function. *st_geometry()* pulls only the geometry column, which can be useful for visualization, and may be required for some layer types or file formats. Our track_point data does not plot without this function.
```{r}
plot(st_geometry(west))
```

For further analysis and prettier mapping with other packages, it may be useful to store the data in a simple tibble. We can extract longitude and latitude from the sf geometry column using the *st_coordinates()* function, which stores longitude (X) and latitude (Y) in a matrix. We will use these to create a tibble containing the time stamp, longitude, and latitude of each track_points.
```{r}
longlat <- st_coordinates(west_sf$geometry) # coordinates extracted from sf geometry column into matrix XY
west2 <- tibble(time = west_sf$time, longitude = longlat[,1], latitude = longlat[,2])
```

## 2. Create maps using {ggmap}
{ggmap} allows us to interact with Google Maps while building maps in a way similar to how we build plots in {ggplot2}. If you want to access Google's satellite maps in R, you must first register with your API key:
```{r}
register_google(key = "AIzaSyDLVTYKWa6uQzmOXgNW5GCCyAyNE2YOa5k")
```
For your own project, replace the key argument in register_google() with your own Maps Static API key, which you can generate by following the instructions here: https://developers.google.com/maps/documentation/javascript/get-api-key
You may then need to fiddle with the settings to ensure that you don't get an error when trying to generate google maps.
1) Under the APIs/Credentials tab, select your key and set "Application restrictions" to None and API restrictions to Restric key, selecting Maps Static API from the dropdown menu. Save.
2) Go to the billing tab and enter billing information. You shouldn't actually be charged unless your API key gets a lot of traffic!

Now that we have access, we can generate a map using the *get_map()* function, which takes arguments including location as the map's center, zoom, and maptype. Here we use satellite - other options include terrain, roadmap, and even watercolor! (https://www.rdocumentation.org/packages/ggmap/versions/3.0.0/topics/get_map)
```{r}
map <- get_map(location = c(lon = 30.400000, lat = 0.489000), zoom = 15, maptype = "satellite")
```
We use *ggmap()* to view the map we've generated, and add other layers using "+", much as we would in {ggplot}.
```{r}
# So this does things but the output doesn't look much like the example?? anyway it's a start
ggmap(map) + geom_density2d(data = west2, aes(x = longitude, y = latitude), colour = "#FFFFFF", size = 0.3) +
  stat_density2d(data = west2, aes(x = longitude, y = latitude, fill = ..level.., alpha = ..level..), size = 0.01, bins = 60, geom = "polygon") +
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0, 0.3), guide = FALSE) +
  geom_point(data = west2, aes(x = longitude, y = latitude,), colour = "#FFFFFF", pch = 16, size = 1)
```

```{r}
# Adapted from code chunk I emailed you guys
# still not sure how it interacts with the map, although I haven't tired grouping by date yet
hulls <- west2 %>%
   #dplyr::group_by(DATE2) %>%
   do(.[chull(.[c('longitude', 'latitude')]), ])
hulls
```

Ended up not getting it to work with get_map, but code for bounding box to define map boundaries
```{r}
bb <- st_bbox(west$geometry)
# ggmap requires it to be stored as list with these variable names:
bb_g <- c(left = bb$xmin, bottom = bb$ymin, right = bb$xmax, top = bb$ymax)
bb_g <- c(left = bb[[1]], bottom = bb[[2]], right = bb[[3]], top = bb[[4]])
```

##Defining home range using minimum convex polygons (MCP) {adehabitat}
##Defining home range using kernels?
##Calculating home range overlap
##Using point-in-polygon to predict group identity?

Random notes for later maybe:
# unclass for st_area() or st_length() which are not traditional vectors but rather have a class of units as metadata.
# big_parks <- parks %>% filter(unclass(areas) > 30000)

# resources
https://r-spatial.github.io/sf/articles/sf1.html