---
title: "ADA-GeospatialVignette-StevenRubenIzzi"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{"ADA-GeospatialVignette-StevenRubenIzzi"}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
This vignette demonstrates how to read in, visualize, and analyze geospatial data. Specifically, we will estimate home ranges and quantify movement using fractal analysis, based on GPS tracks from wild chimpanzees. Let's get started by installing the following packages:
```{r, message=FALSE}
library(GeospatialVignette)
library(sf)
library(sp)
library(tidyverse)
library(ggmap)
library(adehabitatHR)
library(fractaldim)
```
## 1. Read in geospatial data (GPX)
The **{sf}** package allows spatial objects to be stored as data frames that are easily manipulated with {dplyr}. The spatial geometry of a feature is stored in a list-column called geometry, which can be as complex as needed to store more geographic information in a single variable for each feature. The **st_read( )** function can read in a number of different filetypes and automatically recognizes file suffixes. Use **st_drivers( )** to print all file formats that can be read in by st_read( ).

In this example, we will be using the data "PartyTracks_CentralEast" and "PartyTracks_West", which are GPX files containing chimpanzee group movement data collected from January to May 2019 on a handheld GPS tracker by a researcher on foot following the members of two adjacent groups of chimpanzees (Central/East and West) in Kibale National Park, Uganda. **st_layers( )** prints the layer types a file contains, along with the number of features and fields for each.
```{r}
st_layers(PartyTracks_West)
```
The layer we are interested in here is "track_points", which we can specify with the **layer** argument when we read in our files with **st_read( ).**
```{r, message=FALSE}
eastcent.sf <- st_read(PartyTracks_CentralEast, layer = "track_points")
west.sf <- st_read(PartyTracks_West, layer = "track_points")
```
We can do an inital visualization of our data using the **plot** function. **st_geometry( )** pulls only the geometry column, which can be useful for visualization, and may be required for some layer types or file formats. Our track_point data does not plot without this function.
```{r}
plot(st_geometry(west.sf))
```

For further analysis and prettier mapping, it will be useful to extract longitude and latitude from the **{sf}** data frame geometry column using the **st_coordinates( )** function, which stores longitude (X) and latitude (Y) in a matrix. We can stash these in a tibble containing the time stamp, longitude, and latitude of each track_point.
```{r}
# For west
longlat <- st_coordinates(west.sf$geometry) # coordinates extracted from sf geometry column into matrix XY
west <- tibble(time = west.sf$time, longitude = longlat[,1], latitude = longlat[,2])
# For eastcentral
longlat <- st_coordinates(eastcent.sf$geometry)
eastcent <- tibble(time = eastcent.sf$time, longitude = longlat[,1], latitude = longlat[,2])
```

## 2. Defining home ranges using {adehabitatHR} and mapping using {ggmap}
One thing animal movement tracking data can tell us about is home ranges. Here we explore two ways to define chimpanzee home ranges: minimum convex polygons and kernel density estimation. Both can be done with the **{adehabitat}** package - some helpful vignettes can be found here: (https://mran.microsoft.com/snapshot/2017-12-11/web/packages/adehabitatHR/vignettes/adehabitatHR.pdf)

### 2.1 Minimum convex polygons (MCP)
A minimum convex polygon is the smallest polygon around a set of points, where all interior angles are less than 180 degrees. We can create this with the **mcp( )** function from the **{adehabitatHR}** package. First, we must convert our data to a SpatialPointsDataFrame (SPDF), which requires the **{sp}** package's **coordinates( )** function.
```{r}
# New tibble containing id, long, lat
west.sp <- tibble(id = "west", x = west$longitude, y = west$latitude)
# Convert to SPDF by defining coordinates
coordinates(west.sp) <- ~x + y
class(west.sp) # west.sp is now an SPDF!
```
We also need to set the correct coordinate system for our data, which were recorded in the standard WGS84 datum using geographic coordinates (longitude and latitude). UTM (Universal Transverse Mercator) is another coordinate system in which geographic locations are measured by distance (in meters) from the central meridian of the UTM zone in which it lies. Kibale National Park, Uganda, lies in Zone 36N (https://spatialreference.org/ref/epsg/wgs-84-utm-zone-36n/). We we must first tell our SPDF that it's in WGS84 using **proj4string( ),** then transform our longitudes and latitudes to UTM eastings and northings (which are in meters) using **spTransform( ).** Every coordinate system has a standardized unique EPSG code that can be used as shorthand in the **CRS( )** coordinate system function. Read more about coordinate systems and reprojecting vector data in R here:
https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/reproject-vector-data/
```{r}
proj4string(west.sp) <- CRS('+init=epsg:4326') # WGS84
west.sp <- spTransform(west.sp, CRS('+init=epsg:32636')) # UTM zone 36N for Kibale National Park, Uganda 
```
Now we're ready to call **mcp( )** to define our minimum convex polygon. The function takes a **percent** of points to include (vs outliers to exclude), **unin** for units of relocation coordinates ("km" or default "m" - which is why we had to convert from geographic coordinates to UTM), and **unout** for units of output area ("m2" for square meters, "km2" for square kilometers, or default "ha" for hectares).
```{r}
west.mcp <- mcp(west.sp, percent = 95, unout = "km2") # exclude 5% of outlier points. 
west.mcp # 11.85 square km
```

When we include 95% of track points, the west group's home range is 11.85 km^2^. We can visualize this by plotting it against the track points.
```{r, message=FALSE}
plot(west.sp)
plot(west.mcp, col = alpha(1:5, 0.5), add = TRUE) # set transparency with alpha
```

We can also see how the MCP increases in area as we include a greater percentage of points (from 50 to 100 by increments of 5) using **mcp.area( )** and setting the **plotit** argument to true. Note that home range size is in m^2^, not km^2^.
```{r, message = FALSE}
mcp.area(west.sp, percent = seq(50, 100, by = 5), plotit = TRUE)
```

And now we'll repeat some of these steps for the central/east chimpanzee group, but combine the SPDF's of both groups by simply using the **rbind( )** function. **mcp( )** can now automatically use the **id** column to generate a polygon for each group in the data frame.
```{r}
# Create SPDF
eastcent.sp <- tibble(id = "eastcent", x = eastcent$longitude, y = eastcent$latitude)
coordinates(eastcent.sp) <- ~x + y
# Reproject coordinate system to UTM Zone 36N
proj4string(eastcent.sp) <- CRS('+init=epsg:4326')
eastcent.sp <- spTransform(eastcent.sp, CRS('+init=epsg:32636'))
# Combine west and eastcentral SPDF's
alltracks.sp <- rbind(west.sp, eastcent.sp)
```

This time, instead of running mcp( ) and plot( ), we will use a new function **mcp_map( )** included in our **{GeospatialVignette}** package that takes the {sp} dataframe containing id and coordinates, uses the **mcp()** function to print estimated home range areas and create a map of points and associated minimum convex polygons using {ggmap}. In addition to the SpatialPointsDataFrame, it takes an argument *perc* for the percentage of points to include in the polygon (default is 95, meaning 5% of outlier points are excluded). It also takes *units* for the area output (default is "km2", alternatives are "m2" and "ha" for hectares).
```{r, message=FALSE}
mcp_map(sp = alltracks.sp)
```
  
Here is the code for the function:
```{r, eval = FALSE}
mcp_map(sp = alltracks.sp)
mcp_map <- function(sp, perc = 95, units = "km2"){
  # Generate mcp
  sp.mcp <- adehabitatHR::mcp(sp, percent = perc, unout = units)
  # Convert to geographic coordinates
  spgeo <- sp::spTransform(sp, CRS("+proj=longlat"))
  mcpgeo <- sp::spTransform(sp.mcp, CRS("+proj=longlat"))
  # Generate basemap using bounding box from spgeo
  basemap <- ggmap::get_stamenmap(bbox = c(
    left = min(spgeo@coords[,1])-0.005,
    bottom = min(spgeo@coords[,2])-0.005,
    right = max(spgeo@coords[,1])+0.005,
    top = max(spgeo@coords[,2])+0.005),
    zoom = 12)
  # Turn the spatial data frame of points into a regular dataframe for plotting
  spgeo.df <- data.frame(spgeo@coords, id = spgeo@data$id)
  # Map using {ggmap}
  map.mcp <- ggmap::ggmap(basemap) + 
    # "fortify" polygon layer to add geometry to the dataframe
    ggplot2::geom_polygon(data = ggplot2::fortify(mcpgeo), ggplot2::aes(long, lat, colour = id, fill = id), alpha = 0.3) +
    ggplot2::geom_point(data = spgeo.df, ggplot2::aes(x = x, y = y, colour = id), alpha = 0.01, pch = 20)  + # set transparency using alpha and smaller point size use pch.
    labs(x = "Longitude", y = "Latitude")
  map.mcp
  print(sp.mcp)
}
```

The east/central group has a larger home range (17.65 km^2^) than the west group (11.85 km^2^), which makes sense since the west group consists of about 70 chimpanzees, while east/central has about 130!

### 2.2 Kernel density estimation (KDE)
While minimum convex polygons are a quick and easy way to visualize the bounds of animal movement, they tend to overestimate home ranges by including areas that are not actually used. A more accurate method is kernel density estimation, which maps actual utilization distribution (UD) of a habitat. A kernel uses a function to predict how likely use is for each pixel within a grid. The function includes a smoothing factor or bandwidth *h*, which is the distance over which a data point can influence UD - a larger *h* means more smoothing and an increased estimate of home range size.

We can use the **kernelUD( )** function, also included in **{adehabitatHR}**, on our combined SpatialPointsDataFrame to generate kernels for both east/central and west chimpanzee groups, then visualize these using **image( )**. The default *h* is the "reference bandwidth" based on standard deviations of x and y coordinates and total number of relocations or points in our dataset.
```{r}
alltracks.kernel <- kernelUD(alltracks.sp, h = "href") # href = the reference bandwidth
image(alltracks.kernel)
```
  
The kernels alone are difficult to interpret, but we can use our handy dandy **kde_map()** function, which converts them to polygons with the **getverticeshr( )** function from **{adehabitatHR}**. This takes a default **percent** of 95 so that 95% contour lines are used (95% of estimated distribution), as well as units for displaying polygon areas. **kde_map( )** also maps the track points and polygons using **{ggmap}**, much like **mcp_map()**.
```{r, message = FALSE}
kde_map(sp = alltracks.sp)
```
  
And here is what's going on under the hood:
```{r, eval = FALSE}
kde_map <- function(sp, perc = 95, units = "km2"){
  # Generate kernels
  sp.kernels <- adehabitatHR::kernelUD(sp, h = "href")
  # Convert kernels to SpatialPolygons
  sp.kde <- adehabitatHR::getverticeshr(sp.kernels, percent = perc, unout = units)
  # Convert to geographic coordinates
  spgeo <- sp::spTransform(sp, CRS("+proj=longlat"))
  kdegeo <- sp::spTransform(sp.kde, CRS("+proj=longlat"))
  # Generate basemap using bounding box from spgeo
  basemap <- ggmap::get_stamenmap(bbox = c(
    left = min(spgeo@coords[,1])-0.005,
    bottom = min(spgeo@coords[,2])-0.005,
    right = max(spgeo@coords[,1])+0.005,
    top = max(spgeo@coords[,2])+0.005),
    zoom = 12)
  # Turn the spatial data frame of points into a regular dataframe for plotting
  spgeo.df <- data.frame(spgeo@coords, id = spgeo@data$id)
  # Map using {ggmap}
  map.kde <- ggmap::ggmap(basemap) + 
    # "fortify" polygon layer to add geometry to the dataframe
    ggplot2::geom_polygon(data = ggplot2::fortify(kdegeo), ggplot2::aes(long, lat, colour = id, fill = id), alpha = 0.3) +
    ggplot2::geom_point(data = spgeo.df, ggplot2::aes(x = x, y = y, colour = id), alpha = 0.01, pch = 20)  + # set transparency using alpha and smaller point size use pch.
    labs(x = "Longitude", y = "Latitude")
  map.kde
  print(sp.kde)
}
```
Notice that both kernel density estimates of area are lower than those produced by generating minimum convex polygons. East/central has decreased by over 2 km^2^, suggesting they utilize a smaller core area more heavily.

## 4. Fractal analysis using {fractaldim}
Finally, we'll take a look at fractal analysis, which is one method used to quantify animal movement. A fractal analysis will output a fractal dimension (D): a ratio that describes how the details of a pattern change as the scale at which they are measured decreases. As a result, the higher the D of an animal movement pattern, the more complex the pattern is. 

To demonstrate how fractal D can change, we will use the **{fractaldim}** package on the "NgogoTrees" dataset, which is a GPX file containing waypoints of trees the chimpanzees visited to feed, along with a timestamp of when they were first observed feeding at each tree.

```{r}
#Import the data 
feeding <- st_read(NgogoTrees, layer = "waypoints")
```

Let's split up the data by time, so we can compare trees chimps visited at three different time points.
```{r}
#Early timepoint
early <- feeding %>%
  filter(time <= as.Date("2019-03-15"))
earlylong <- sapply(early$geometry,"[[", 2)
earlylat <- sapply(early$geometry,"[[", 1)

#Middle timepoint
mid <- feeding %>%
  filter(time >= as.Date("2019-03-15") & time <= as.Date("2019-05-01"))
midlong <- sapply(mid$geometry,"[[", 2)
midlat <- sapply(mid$geometry,"[[", 1)

#Late timepoint
late <- feeding%>%
  filter(time >= as.Date("2019-05-01"))
latelong <- sapply(late$geometry,"[[", 2)
latelat <- sapply(late$geometry,"[[", 1)
```

Now we will use a function called **fractal.map( )** from our {GeospatialVignette} package, which makes a simple plot of the coordinates using ggplot, then uses the **fd.estim.dctII( )** function from **{fractaldim}** to conduct a fractal analysis and compute a D value. We will run this on each of the three timepoints. 
```{r, message = FALSE}
fractal_plot(early,earlylat,earlylong)
fractal_plot(mid,midlat,midlong)
fractal_plot(late,latelat,latelong)
```
  
The three timepoints vary slightly in terms of Fractal D, indicating that there may be a difference in how the animals are visiting their feeding trees over time. Given the middle timepoint had the lowest fractal D, it is possible that at that time of year the animals may be moving around from tree to tree less than at the other two times.

And here is the code for the **fractal.plot()** function:
```{r, eval = FALSE}
fractal_plot <- function(d,latitude,longitude) {
  plot1 <- ggplot2::ggplot(data = d,  ggplot2::aes(latitude,longitude)) + ggplot2::geom_point()
  fractalplot <- fractaldim::fd.estim.dctII(cbind(latitude,longitude),plot.loglog=TRUE, plot.allpoints=TRUE, nlags="auto")
  return(list(plot1, fractalplot))
}
```

## Conclusion
We've provided some examples of how to read in GPX data using **{sf}**, estimate animal home ranges using **{sp}** objects and **{adehabitatHR}**, map geospatial data using **{ggmap}**, and use fractal analysis to analyze space use patterns with **{fractaldim}**, but we've just barely scratched the surface of what can be done with geospatial data in R! Here are some more resources:  
  
https://rspatial.org/  
https://geocompr.robinlovelace.net/  
https://r-spatial.github.io/sf/articles/sf1.html
https://jamesepaterson.github.io/jamespatersonblog/01_trackingworkshop_formatting (our ggmap code is adapted from here)